---
title: "Analysis 5: : Morphological differentiation analysis"
author: "Joe Gunn"
date: "2022-07-28"
output: html_document
---

# Project: Population genetic structure and morphological differentiation between Northern Smallmouth Bass (<i>Micropterus dolomieu dolomieu</i>) and Neosho Smallmouth Bass (<i>M. d. velox</i>)
We investigated patterns of genetic (via microsatellites) and morphological diversity, differentiation, and structure across Smallmouth Bass populations (<i>Micropterus dolomieu</i>) in the Central Interior Highlands (CIH) of North America, including the two recognized subspecies: Northern Smallmouth Bass (<i>M. d. dolomieu</i>), which is native to the lower Ozark Highlands, and Neosho Smallmouth Bass (<i>M. d. velox</i>), which is endemic to tributaries of the Arkansas River Basin. We compared three independent combinations of starting parameters in the Bayesian clustering software program STRUCTURE to ascertain a robust picture of hierarchical genetic structure in the CIH. We paired these data with genetic diversity metrics (e.g., observed and expected heterozygosity and allelic richness) to determine relative amounts of variation across geographically separated populations. We then assessed differentiation in five morphometric and meristic traits. We ultimately aimed to validate or amend the taxonomic status of the Smallmouth Bass subspecies by revealing potential ecological and evolutionary divergence between them, with the hope that increased taxonomic resolution would provide insight into the presence and distribution of evolutionary significant and management units for a popular sportfish.

## Specific Aim: Genetic diversity and differentiation analysis
In this analysis, we assess morphological differentiation between Neosho Smallmouth Bass and Northern Smallmouth Bass in the CIH using five morphometrics (traits with continuous distributions; total length, standard length, body depth, head length, and orbital length) and one meristic (discrete count trait; soft dorsal fin rays) and three independently analysis frameworks: 1) individual analyses of covariance (ANCOVAs) on morphometric traits; 2) multiviariate discriminant function analysis (DFA) on morphometric traits; and 3) logistic regression on soft dorsal fin rays. With ANCOVA, we assess differentiation between subsepecies at each trait individually. With DFA, we assess the accuracy with which an individual can be assigned to either subspecies based on composite trait values. With logistic regression, we assess the probability with which a Neosho Smallmouth Bass and Northern Smallmouth Bass possess a specific number of soft dorsal rays.

## Phases of Analysis
### Phase 1: Univariate morphometric analysis with ANCOVA
### Phase 2: Multivariate morphometric analysis with DFA
### Phase 3: Univariate meristic analysis with logistic regression

## Libraries needed for analysis
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(pophelper)
library(ggpubr)
library(nlme)
```

## PHASE 1: UNIVARIATE MORPHOMETRIC ANALYSIS WITH ANCOVA
In this phase of the analysis, we assess differentiation between Neosho Smallmouth Bass and Northern Smallmouth Bass in four individual morphometric traits (i.e., standard length, body depth, head length, and orbital length; see Analysis 2 in `../02_data_summary_analysis/smb_genetics_data_summary_analysis` for descriptions for each trait) using analysis of covariance, comparing each trait between subspecies as they co-vary with total length. We complete this assessment for two subsets of samples: 1) all fish labeled by the subspecies range from which they were collected (Arkansas River Basin, aka Neosho, vs. Missouri or White River drainages, aka Northern), and 2) only fish that assigned as purely (defined as assigning with >90% ancestry membership to any one genetic cluster based on population structure analysis) Neosho or Northern. 

### STEP 1: Read in and combine raw phenotype data, metadata, and population structure membership data; run the Rmd chunk below
In this step, we prepare the data for analysis by merging all phenotype data with both metadata and population structure membership data as ascertained in Analysis 3.

##### Read in phenotype data and merge with metadata:
```{r}
# read in phenotpye data
load("../02_data_summary_analysis/data/phenotype_data.Rda")

# load sample metadata
load("../02_data_summary_analysis/data/metadata.Rda")

# Read in combined-merged Q table for K=3 from default structure analysis
q <- readQ("../03_structure_analysis/data/structure_data/clumpp_data/default_clumpp/full_data/pop_K3/pop_K3-combined-merged.txt") %>%
  as.data.frame()

# Combine phenotype data and metadata
phenotypes <- full_join(metadata, 
                        phenotype_data,
                        by = "sample_id")

# Get metadata for only genotyped samples
metadata_genotypes <- metadata %>%
  filter(genotype == "yes")

# Combine structure data and metadata
q <- cbind(metadata_genotypes$sample_id, 
           q)

# rename columns by predominant cluster membership 
colnames(q) <- c("sample_id", "spotted_bass", "neosho_smallmouth_bass", "northern_smallmouth_bass")

# Bind genotyped sample_id column with population structure membership data
phenotypes <- full_join(phenotypes,
                        q,
                        by = "sample_id")

# Eliminate spotted bass from the dataset altogether
phenotypes <- phenotypes %>%
  filter(subspecies != "Spotted Bass") %>%
  dplyr::select(-spotted_bass)

# Save full phenotype data for downstream analyses
save(phenotypes, file = "data/phenotypes.Rda")
```

### STEP 2: Run ANCOVA on individual morphometric traits
In this step, we run analysis of covariance (ANCOVA) on four morphometric traits (standard length, SL; body depth, BD; head length, HL; and orbital length, OL) individually. We regress each trait on TL and compare regressions between subspecies by range (where samples were collected) and by genetic assignment (samples assigning > 90% to either Neosho or Northern Smallmouth Bass genetic clusters as inferred in population structure analysis).

#### 2a. Generate separate datasets for 1) morphometric traits by range and 2) morphometric traits by genetic assignment and summarize sample sizes and log-transform trait values; run the rmd chunk below.

##### Read in phenotype data, merge with metadata, and generate range and genetics datasets with log-transformed trait values:
```{r}
# read in phenotpye data
load("data/phenotypes.Rda")

# Generate and clean dataset for samples by range
range <- phenotypes %>%
  dplyr::select(sample_id, population, subspecies, tl, sl, ol, hl, bd) %>%
  drop_na()

# Generate and clean dataset for samples by genetic assignment
genetics <- phenotypes %>%
  dplyr::select(sample_id, population, tl, sl, ol, hl, bd, northern_smallmouth_bass, neosho_smallmouth_bass) %>%
  filter(northern_smallmouth_bass > 0.9 |
           neosho_smallmouth_bass > 0.9) %>%
  drop_na()

# Categorize sample ids into 25 mm size bins based on total length while alive (tl_alive)
genetics$genetic_id <- with(genetics, ifelse(northern_smallmouth_bass >= 0.9, "Northern Smallmouth Bass",
                                             ifelse(northern_smallmouth_bass <= 0.1, "Neosho Smallmouth Bass", "done")))

# Log transform trait values in range data 
range <- range %>%
  mutate(logtl = log(tl),
         logsl = log(sl),
         logbd = log(bd),
         loghl = log(hl),
         logol = log(ol))

# Log transform trait values in genetics data 
genetics <- genetics %>%
  mutate(logtl = log(tl),
         logsl = log(sl),
         logbd = log(bd),
         loghl = log(hl),
         logol = log(ol))

# Count samples by range
range %>%
  count()

# Count samples by subspecies and range
range %>%
  group_by(subspecies) %>%
   count()

# Count samples by genetic id
genetics %>%
  count()

# Count samples by subspecies and range
genetics %>%
  group_by(genetic_id) %>%
   count()

# Save range data
save(range, file = "data/range.Rda")

# Save genetics data
save(genetics, file = "data/genetics.Rda")
```

### Data summary:

## Total samples by range
<b><i>N</i><sub>range</sub></b> = 249 <br>

## Total samples by subspecies and range
<b><i>N</i><sub>Neosho Smallmouth Bass </sub></b> = 184 <br>
<b><i>N</i><sub>Northern Smallmouth Bass</sub></b> = 65 <br>

## Total samples by range
<b><i>N</i><sub>range</sub></b> = 89 <br>

## Total samples by subspecies and range
<b><i>N</i><sub>Neosho Smallmouth Bass </sub></b> = 17 <br>
<b><i>N</i><sub>Northern Smallmouth Bass</sub></b> = 72 <br>

#### 2b. Test normality in all morphometric traits

##### 2b.1. Test for normality in Neosho Smallmouth Bass samples by range; run the Rmd chunk below.

##### Test for normality in trait values:
```{r}
# Load range data
load("data/range.Rda")

# Get only neosho samples by range
neosho_range <- range %>%
  filter(subspecies == "Neosho Smallmouth Bass")

# Test for normality in each trait
shapiro.test(neosho_range$logtl)
shapiro.test(neosho_range$logsl)
shapiro.test(neosho_range$logol)
shapiro.test(neosho_range$loghl)
shapiro.test(neosho_range$logbd)
```

<b> Model results by trait</b>: <br>
TL: p = 0.07977, W = 0.98668 (Normal)
SL: p = 0.09018, W = 0.98707 (Normal)
OL: p < 0.00100, W = 0.95523 (Non-normal)
HL: p = 0.16490, W = 0.988970 (Normal)
BD: p = 0.59970, W = 0.99355 (Normal)

##### 2b.2. Test for normality in Northern Smallmouth Bass samples by range; run the Rmd chunk below.

##### Test for normality in trait values:
```{r}
# Load range data
load("data/range.Rda")

# Get only northern samples by range
northern_range <- range %>%
  filter(subspecies == "Northern Smallmouth Bass")

# Test for normality in each trait
shapiro.test(northern_range$logtl) 
shapiro.test(northern_range$logsl) 
shapiro.test(northern_range$logol)
shapiro.test(northern_range$loghl)
shapiro.test(northern_range$logbd) 
```

<b> Model results by trait</b>: <br>
TL: p = 0.4577, W = 0.98188 (Normal)
SL: p = 0.2968, W = 0.97793 (Normal)
OL: p = 0.5188, W = 0.98312 (Normal)
HL: p < 0.0010, W = 0.91592 (Non-normal)
BD: p = 0.2971, W = 0.97794 (Normal)

##### 2b.3. Test for normality in Neosho Smallmouth Bass samples by genetic assignment; run the Rmd chunk below.

##### Test for normality in trait values:
```{r}
# Load genetics data
load("data/genetics.Rda")

# Get only neosho samples by genetic assignment
neosho_genetics <- genetics %>%
  filter(genetic_id == "Neosho Smallmouth Bass")

# Test for normality in each trait
shapiro.test(neosho_genetics$logtl)
shapiro.test(neosho_genetics$logsl)
shapiro.test(neosho_genetics$logol)
shapiro.test(neosho_genetics$loghl)
shapiro.test(neosho_genetics$logbd)
```

<b> Model results by trait</b>: <br>
TL: p = 0.4462, W = 0.94934 (Normal)
SL: p = 0.5208, W = 0.95389 (Normal)
OL: p = 0.5098, W = 0.95324 (Normal)
HL: p = 0.6847, W = 0.96281 (Normal)
BD: p = 0.2028, W = 0.92817 (Normal)

##### 2b.4. Test for normality in Northern Smallmouth Bass samples by genetic assignment; run the Rmd chunk below.

##### Test for normality in trait values:
```{r}
# Load genetics data
load("data/genetics.Rda")

# Get only neosho samples by genetic assignment
northern_genetics <- genetics %>%
  filter(genetic_id == "Northern Smallmouth Bass")

# Test for normality in each trait
shapiro.test(northern_genetics$logtl) 
shapiro.test(northern_genetics$logsl) 
shapiro.test(northern_genetics$logol) 
shapiro.test(northern_genetics$loghl)
shapiro.test(northern_genetics$logbd) 
```

<b> Model results by trait</b>: <br>
TL: p = 0.6208, W = 0.98621 (Normal)
SL: p = 0.7938, W = 0.98910 (Normal)
OL: p = 0.5048, W = 0.98421 (Normal)
HL: p < 0.0010, W = 0.90931 (Non-normal)
BD: p = 0.6511, W = 0.98671 (Normal)

##### 2b.5. Plot qq normal distribution plots for all traits and groups; run the Rmd chunk below.

##### Plot qq normal distribution for all traits: `figures/normality.pdf`
```{r}
# Load range data
load("data/range.Rda")

# Load genetics data
load("data/genetics.Rda")

# Get only neosho samples by range
neosho_range <- range %>%
  filter(subspecies == "Neosho Smallmouth Bass")

# Neosho range total length plot
neosho_range_tl <- ggqqplot(neosho_range$logtl) + 
    labs(y = "Total Length Sample", title = "Neosho Range") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(axis.title.y = element_text(size = 20)) + 
    theme(plot.title = element_text(size = 20)) + 
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho range standard length plot
neosho_range_sl <- ggqqplot(neosho_range$logsl) + 
    labs(y = "Standard Length Sample") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_text(size = 20)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho range orbital length plot
neosho_range_ol <- ggqqplot(neosho_range$logol) + 
    labs(y = "Orbital Length Sample") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_text(size = 20)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho range head length plot
neosho_range_hl <- ggqqplot(neosho_range$loghl) + 
    labs(y = "Head Length Sample") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_text(size = 20))  +
    theme(panel.border = element_rect(colour =  "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho range body depth plot
neosho_range_bd <- ggqqplot(neosho_range$logbd) + 
    labs(y = "Body Depth Sample", x = "Theoretical") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.y = element_text(size = 20)) + 
    theme(axis.title.x = element_text(size = 20))  +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Get only neosho samples by range
northern_range <- range %>%
  filter(subspecies == "Northern Smallmouth Bass")

# Northern range total length plot
northern_range_tl <- ggqqplot(northern_range$logtl) + 
    labs(title = "Northern Range") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(plot.title = element_text(size = 20))  +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern range standard length plot
northern_range_sl <- ggqqplot(northern_range$logsl) + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) +
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern range orbital length plot
northern_range_ol <- ggqqplot(northern_range$logol) +  
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern range head length plot
northern_range_hl <- ggqqplot(northern_range$loghl) + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern range body depth plot
northern_range_bd <- ggqqplot(northern_range$logbd) + 
    labs(x = "Theoretical") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_text(size = 20)) + 
    theme(axis.title.y = element_blank())  +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Get only Neosho samples by genetic assignment
neosho_genetics <- genetics %>%
  filter(genetic_id == "Neosho Smallmouth Bass")

# Neosho genetics total length plot
neosho_genetics_tl <- ggqqplot(neosho_genetics$logtl) + 
    labs(title = "Neosho Genetic Assignment") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(plot.title = element_text(size = 20)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho genetics standard length plot
neosho_genetics_sl <- ggqqplot(neosho_genetics$logsl) + 
    theme_set(theme_cowplot(12)) +
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho genetics orbital length plot
neosho_genetics_ol <- ggqqplot(neosho_genetics$logol) + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho genetics head length plot
neosho_genetics_hl <- ggqqplot(neosho_genetics$loghl) + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Neosho genetics body depth plot
neosho_genetics_bd <- ggqqplot(neosho_genetics$logbd) + 
    labs(x = "Theoretical") + theme_set(theme_cowplot(12)) +
    theme(axis.title.y = element_blank()) +
    theme(axis.title.x = element_text(size = 20)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Get only Northern samples by genetic assignment
northern_genetics <- genetics %>%
  filter(genetic_id == "Northern Smallmouth Bass")

# Northern genetics total length plot
northern_genetics_tl <- ggqqplot(northern_genetics$logtl) + 
    labs(title = "Northern Genetic Assignment") + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(plot.title = element_text(size = 20)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern genetics standard length plot
northern_genetics_sl <- ggqqplot(northern_genetics$logsl) + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern genetics orbital length plot
northern_genetics_ol <- ggqqplot(northern_genetics$logol) +
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) +
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern genetics head length plot
northern_genetics_hl <- ggqqplot(northern_genetics$loghl) + 
    theme_set(theme_cowplot(12)) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Northern genetics body depth plot
northern_genetics_bd <- ggqqplot(northern_genetics$logbd) + 
    theme_set(theme_cowplot(12)) + 
    labs(x = "Theoretical") + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_text(size = 20)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0))

# Plot all together
pdf("figures/qq_plots.pdf", width = 19, height = 19)

plot_grid(neosho_range_tl, northern_range_tl, neosho_genetics_tl, northern_genetics_tl, 
          neosho_range_sl, northern_range_sl, neosho_genetics_sl, northern_genetics_sl,
          neosho_range_ol, northern_range_ol, neosho_genetics_ol, northern_genetics_ol,
          neosho_range_hl, northern_range_hl, neosho_genetics_hl, northern_genetics_hl,
          neosho_range_bd, northern_range_bd, neosho_genetics_bd, northern_genetics_bd,
          ncol = 4,
          nrow = 5,
          labels = c("a","b","c","d","e",
                     "f","g","h","i","j",
                     "k","l","m","n","o",
                     "p","q","r","s","t"),
          label_size = 25,
          label_y = 1.03)

dev.off()
```

This figure is the basis for Figure S4 in the final ms.

#### 2c. Plot and compare density distributions for all traits by range and genetic assignment; run the Rmd chunk below.

##### Plot denisty distributions for all traits: `figures/density.pdf`
```{r}
# Load range data
load("data/range.Rda")

# Load genetics data
load("data/genetics.Rda")

# Change names of subspecies for plotting
levels(range$subspecies) <- c("Neosho", "Northern", "Spotted")

# Range total length plot
range_tl <- ggplot(range, aes(x = logtl, fill = subspecies)) + 
    geom_density(show.legend = T, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    labs(fill = "Range", y = "Probability Density", fill = "Subspecies") + 
    theme(legend.position = c(0.6,0.85), legend.background = element_blank()) + 
    theme(legend.text = element_text(size = 15), legend.title = element_text(size = 20, face = "bold")) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.title.y = element_text(size = 20)) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Range standard length plot
range_sl <- ggplot(range, aes(x = logsl, fill = subspecies)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Range orbital length plot
range_ol <- ggplot(range, aes(x = logol, fill = subspecies)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.text = element_text(size = 20)) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Range head length plot
range_hl <- ggplot(range, aes(x = loghl, fill = subspecies)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_blank()) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Range body depth plot
range_bd <- ggplot(range, aes(x = logbd, fill = subspecies)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    theme(axis.title.x = element_blank()) + theme(axis.title.y = element_blank()) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Plot all range together
range <- plot_grid(range_tl, 
                   range_sl, 
                   range_ol, 
                   range_hl, 
                   range_bd, 
                   nrow = 1,
                   labels = c("f","g","h","i","j"),
                   label_size = 25,
                   label_y = 0.99,
                   label_x = 0.1)

# Genetics total length plot
genetics_tl <- ggplot(genetics, aes(x = logtl, fill = genetic_id)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    labs(fill = "Assignment", y = "Probability Density", x = "Total Length (mm)") + 
    theme(legend.position = c(0.6,0.9), legend.background = element_blank()) + 
    theme(legend.text = element_text(size = 15), legend.title = element_text(size = 20, face = "bold")) + 
    theme(axis.title = element_text(size = 20)) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Genetics standard length plot
genetics_sl <- ggplot(genetics, aes(x = logsl, fill = genetic_id)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    labs(x = "Standard Length (mm)") + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_text(size = 20)) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Genetics orbital length plot
genetics_ol <- ggplot(genetics, aes(x = logol, fill = genetic_id)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    labs(x = "Orbital Length (mm)") + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_text(size = 20)) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Genetics head length plot
genetics_hl <- ggplot(genetics, aes(x = loghl, fill = genetic_id)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    labs(x = "Head Length (mm)") + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_text(size = 20)) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Genetics body depth plot
genetics_bd <- ggplot(genetics, aes(x = logbd, fill = genetic_id)) + 
    geom_density(show.legend = F, alpha = 0.7) + 
    theme_set(theme_cowplot(12)) + 
    scale_fill_manual(values = c("springgreen3", "mediumpurple")) + 
    labs(x = "Body Depth (mm)") + 
    theme(axis.title.y = element_blank()) + 
    theme(axis.title.x = element_text(size = 20)) + 
    theme(axis.text = element_text(size = 20)) +
    scale_x_continuous(expand = c(0,0)) + 
    scale_y_continuous(expand = c(0,0)) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

# Plot all genetics together
genetics <- plot_grid(genetics_tl, 
                      genetics_sl, 
                      genetics_ol, 
                      genetics_hl, 
                      genetics_bd, 
                      nrow = 1,
                      labels = c("a","b","c","d","e"),
                      label_size = 25,
                      label_y = 0.99,
                      label_x = 0.01)

# Plot everything together
pdf("figures/density.pdf", width = 20, height = 8)

plot_grid(range, 
          genetics,
          nrow = 2)

dev.off()
```

This figure is the basis for Figure S5 in the final ms.

#### 2d. Run ANCOVA models for all traits

##### 2d.1. Run ANCOVA models for traits by range

###### 2d.1.1. Run ANCOVAs with interaction terms and extract model coefficients by range; run the Rmd chunk below

##### Run ANCOVAs including interaction term:
```{r}
# Load range data
load("data/range.Rda")

# ANCOVA with interaction for standard length
range_sl <- lme(logsl ~ logtl*subspecies, 
                random = ~1|population,
                data = range)

# Get summary statistics
anova(range_sl)

# ANCOVA with interaction for orbital length
range_ol <- lme(logol ~ logtl*subspecies,
                random = ~1|population, 
                data = range)

# Get summary statistics
anova(range_ol)

# ANCOVA with interaction for head length
range_hl <- lme(loghl ~ logtl*subspecies, 
                random = ~1|population, 
                data = range)

# Get summary statistics
anova(range_hl)

# ANCOVA with interaction for body depth
range_bd <- lme(logbd ~ logtl*subspecies, 
                random = ~1|population, 
                data = range)

# Get summary statistics
anova(range_bd)
```

<b> Interaction model results by trait</b>: <br>
SL: p = 0.1411 , F = 2.2000 (no interaction)
OL: p < 0.0001 , F = 33.301 (interaction)
HL: p = 0.8227 , F = 0.0500 (no interaction)
BD: p = 0.2085 , F = 1.5900 (no interaction)

###### 2d.1.2. Run additive ANCOVAs and extract model coefficients by range; run the Rmd chunk below

##### Run additive ANCOVAs models:
```{r}
# Load range data
load("data/range.Rda")

# ANCOVA with interaction for standard length
range_sl <- lme(logsl ~ logtl + subspecies, 
                random = ~1|population,
                data = range)

# Get summary statistics
anova(range_sl)

# ANCOVA with interaction for orbital length
range_ol <- lme(logol ~ logtl + subspecies,
                random = ~1|population, 
                data = range)

# Get summary statistics
anova(range_ol)

# ANCOVA with interaction for head length
range_hl <- lme(loghl ~ logtl + subspecies, 
                random = ~1|population, 
                data = range)

# Get summary statistics
anova(range_hl)

# ANCOVA with interaction for body depth
range_bd <- lme(logbd ~ logtl + subspecies, 
                random = ~1|population, 
                data = range)

# Get summary statistics
anova(range_bd)
```

<b> Model results by trait</b>: <br>
SL: p = 0.2341, F = 1.600 (not significant)
OL: p = 0.5231, F = 0.431 (not significant)
HL: p = 0.0067, F = 10.37 (significant)
BD: p = 0.8468, F = 0.040 (not significant)

##### 2d.2. Run ANCOVA models for traits by genetic assignment

###### 2d.2.1. Run ANCOVAs with interaction terms and extract model coefficients by genetic assignment; run the Rmd chunk below

##### Run ANCOVAs including interaction term:
```{r}
# Load genetics data
load("data/genetics.Rda")

# ANCOVA with interaction for standard length
genetics_sl <- lme(logsl ~ logtl*genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_sl)

# ANCOVA with interaction for orbital length
genetics_ol <- lme(logol ~ logtl*genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_ol)

# ANCOVA with interaction for head length
genetics_hl <- lme(loghl ~ logtl*genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_hl)

# ANCOVA with interaction for body depth
genetics_bd <- lme(logbd ~ logtl*genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_bd)
```

<b> Interaction model results by trait</b>: <br>
SL: p = 0.1111 , F = 2.6000 (no interaction)
OL: p = 0.1129 , F = 2.5740 (no interaction)
HL: p = 0.7336 , F = 0.1200 (no interaction)
BD: p = 0.6771 , F = 0.1700 (no interaction)

###### 2d.2.2 Run additive ANCOVAs and extract model coefficients by genetic assignment; run the Rmd chunk below

##### Run additive ANCOVAs models:
```{r}
# Load genetics data
load("data/genetics.Rda")

# ANCOVA with interaction for standard length
genetics_sl <- lme(logsl ~ logtl + genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_sl)

# ANCOVA with interaction for orbital length
genetics_ol <- lme(logol ~ logtl + genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_ol)

# ANCOVA with interaction for head length
genetics_hl <- lme(loghl ~ logtl + genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_hl)

# ANCOVA with interaction for body depth
genetics_bd <- lme(logbd ~ logtl + genetic_id, 
                   random = ~1|population, 
                   data = genetics)

# Get summary statistics
anova(genetics_bd)
```

<b> Model results by trait</b>: <br>
SL: p = 0.5694, F = 0.300 (not significant)
OL: p = 0.9160, F = 0.011 (not significant)
HL: p = 0.1131, F = 2.570 (not significant)
BD: p = 0.2706, F = 1.230 (not significant)

## --------------  END OF ANALYSIS 5: GENETIC DIVERSITY AND DIFFERENTIATION ANALYSIS --------------  ##

