---
title: "NSMB_Microsats_GENETIC_DIVERSITY"
author: "Joe Gunn"
date: "7/25/2019"
output: html_document
---

# Libraries Needed For Analysis
```{r Libraries, echo = FALSE, include = FALSE}

library(riverdist)
library(sp)
library(readxl)
library(popgraph)
library(tidyverse)
library(gstudio)
library(ade4)
library(adegenet)
library(cowplot)
library(nlme)
library(poppr)
library(PopGenReport)
library(maps)
library(mapdata)
library(stringr)
library(rgdal)
library(sf)
library(ggsn)
library(raster)
library(lme4)
library(factoextra)
library(MASS)
library(logihist)
library(pophelper)
library(ggpubr)
library(DescTools)
library(vegan)
library(ggstatsplot)
```

## Metadata
```{r}
smb_metadata_geo <- read_excel("../../data/metadata/smb_metadata_geo.xlsx")
smb_metadata_geo_nor <- smb_metadata_geo[c(617:766),]
smb_metadata_neo <- read_excel("../../data/metadata/smb_metadata_nonGEO.xlsx")
```

## Read in genetic data and clean data.
```{r}
gen_data <- read_population("../../data/microsatellite_data//micros_blank_river.csv", type = "column", locus.columns = 6:33)

#Data Cleaning:
#Change categorical variables to factors
gen_data$Population_Name <- as.factor(gen_data$Population_Name)
gen_data$Sample_ID <- as.factor(gen_data$Sample_ID)
```

#Calculate Genetic Diversity Metrics
```{r}
#Common Allele Frequency
A_95 <- genetic_diversity(gen_data, mode = "A95") # frequency of non-rare alleles by locus

A_95_bypop <- genetic_diversity(gen_data, mode = "A95", stratum = "Population_Name") # non-rare allele frequency by individual
A_95_bypop <- A_95_bypop %>% 
  group_by(Stratum) %>% 
  summarize (mean_A95 = mean(A95)) # average non-rare allele frequency by population

#Heterozygosity
Ho_pop <- genetic_diversity(gen_data, stratum = "Population_Name", mode = c("Ho")) # calculate observed heterozygosity across populations 
He_pop <- genetic_diversity(gen_data, stratum = "Population_Name", mode = c("He")) # calculate expected heterozygosity across populations

Ho_pop_ave <- Ho_pop %>% 
  group_by(Stratum) %>% 
  summarize(mean_Ho = mean(Ho)) # calculate average observed heterozygosity for populations

He_pop_ave <- He_pop %>% 
  group_by(Stratum) %>% 
  summarize(mean_He = mean(He)) # calculate average expected heterozygosity for populations
```

# Read-in new data file compiled in Excel and clean dataframe to compute averages of each metric by collection site. All of the information in the dataframe imported below were calculated in the above section (Calculate Genetic Diversity Metrics) or in outside software programs not built into R.
```{r}
#read-in new data file compiled in Excel from data frames above
options(scipen = 999)
div <- read_excel("../../data/excel_data/div_CERC.xls")

#Clean and organize data files by subspecies range
div_ave <- div %>% 
  group_by(Range) %>% 
  summarize(meanAR = mean(AR), meanAE = mean(AE), mean95 = mean(A95), meanAP = mean(AP), meanHO = mean(HO), meanHE = mean(HE)) # calculate mean of diversity metrics at subsepecies level

div_sd <- div %>% 
  group_by(Range) %>% 
  summarize(sdAR = sd(AR), sdAE = sd(AE), sd95 = sd(A95), sdAP = sd(AP), sdHO = sd(HO), sdHE = sd(HE)) # calculate standard deviation of diversity metrics at subspecies level
div_nospb <- div[-c(41:43),]

div_neo <- div %>% 
  filter(Range == "Neosho") #create data file for Neosho samples only

div_nor <- div %>% 
  filter(Range == "Northern") #create data file for Northern only

div_spb <- div %>% 
  filter(Range == "Spotted") #create data file for Spotted only

div_neo_ave <- div_neo %>% 
  group_by(Range) %>% 
  summarize(AR = mean(AR), A95 = mean(A95), AP = mean(AP), HO = mean(HO), HE = mean(HE))

div_nor_ave <- div_nor %>% 
  group_by(Range) %>% 
  summarize(AR = mean(AR), A95 = mean(A95), AP = mean(AP), HO = mean(HO), HE = mean(HE))

div_neo_sd <- div_neo %>% 
  summarize(AR = sd(AR), A95 = sd(A95), AP = sd(AP), HO = sd(HO), HE = sd(HE))

div_nor_sd <- div_nor %>% 
  summarize(AR = sd(AR), A95 = sd(A95), AP = sd(AP), HO = sd(HO), HE = sd(HE))

div_neo_sd <- t(div_neo_sd)
div_neo_sd <- as.data.frame(div_neo_sd)
div_neo_sd <- div_neo_sd %>% 
  rownames_to_column("metric")
colnames(div_neo_sd) <- c("metric", "sd")

div_neo_ave <- t(div_neo_ave)
div_neo_ave <- as.data.frame(div_neo_ave)
div_neo_ave <- div_neo_ave %>% 
  rownames_to_column("metric")
colnames(div_neo_ave) <- c("metric", "mean")
div_neo_ave <- div_neo_ave[-1,]

div_nor_sd <- t(div_nor_sd)
div_nor_sd <- as.data.frame(div_nor_sd)
div_nor_sd <- div_nor_sd %>% 
  rownames_to_column("metric")
colnames(div_nor_sd) <- c("metric", "sd")

div_nor_ave <- t(div_nor_ave)
div_nor_ave <- as.data.frame(div_nor_ave)
div_nor_ave <- div_nor_ave %>% 
  rownames_to_column("metric")
colnames(div_nor_ave) <- c("metric", "mean")
div_nor_ave <- div_nor_ave[-1,]

neo_metrics <- merge(div_neo_ave, div_neo_sd, by = "metric")
nor_metrics <- merge(div_nor_ave, div_nor_sd, by = "metric")
nor_labels <- c(rep("Northern", 5))
nor_labels <- as.data.frame(nor_labels)
colnames(nor_labels) <- c("range")

neo_labels <- c(rep("Neosho", 5))
neo_labels <- as.data.frame(neo_labels)
colnames(neo_labels) <- c("range")

neo_metrics <- cbind(neo_labels, neo_metrics)
nor_metrics <- cbind(nor_labels, nor_metrics)
genetic_metrics <- rbind(neo_metrics, nor_metrics)
```

## Run linear models to test for significant differences in genetic diversity at the subspecies level
```{r}
lm1 <- aov(HO ~ Range, data = div_nospb) # observed heterozygosity
summary(lm1)

lm2 <- lm(HE ~ Range, data = div_nospb) # expected heterozygosity
summary(lm2)

lm3 <- lm(AR ~ Range, data = div_nospb) # rarefied allelic richness (taken from rarefied allelic richness run in HP-Rare)
summary(lm3)

lm5 <- lm(A95 ~ Range, data = div_nospb) # non-rare allele frequency
summary(lm5)

lm6 <- lm(AP ~ Range, data = div_nospb) # private (rare) allele frequency (taken from private allele frequency run in HP-Rare)
summary(lm6)
```

## Calculate 95 % confidence intervals for each metric
```{r}
e95neo <- qt(0.975,df=30)*0.82168581/sqrt(30)
eAPneo <- qt(0.975,df=30)*0.03961101/sqrt(30)
eARneo <- qt(0.975,df=30)*0.57309329/sqrt(30)
eHEneo <- qt(0.975,df=30)*0.09935171/sqrt(30)
eHOneo <- qt(0.975,df=30)*0.09809077/sqrt(30)
e95nor <- qt(0.975,df=8)*0.91145228/sqrt(8)
eAPnor <- qt(0.975,df=8)*0.05897269/sqrt(8)
eARnor <- qt(0.975,df=8)*0.48365794/sqrt(8)
eHEnor <- qt(0.975,df=8)*0.06055186/sqrt(8)
eHOnor <- qt(0.975,df=8)*0.05367262/sqrt(8)

error <- c(e95neo, eAPneo, eARneo, eHEneo, eHOneo, e95nor, eAPnor, eARnor, eHEnor, eHOnor)
error <- as.data.frame(error)
genetic_metrics <- cbind(genetic_metrics, error)

A95_data <- genetic_metrics %>% 
  filter(metric == "A95")

A95_data <- A95_data %>% 
  mutate(mean = as.character(mean)) %>% 
  mutate(mean = as.numeric(mean))

AP_data <- genetic_metrics %>% 
  filter(metric == "AP")

AP_data <- AP_data %>% 
  mutate(mean = as.character(mean)) %>% 
  mutate(mean = as.numeric(mean))

AR_data <- genetic_metrics %>% 
  filter(metric == "AR")

AR_data <- AR_data %>% 
  mutate(mean = as.character(mean)) %>% 
  mutate(mean = as.numeric(mean))

HO_data <- genetic_metrics %>% 
  filter(metric == "HO")

HO_data <- HO_data %>% 
  mutate(mean = as.character(mean)) %>% 
  mutate(mean = as.numeric(mean))

HE_data <- genetic_metrics %>% 
  filter(metric == "HE")

HE_data <- HE_data %>% 
  mutate(mean = as.character(mean)) %>% 
  mutate(mean = as.numeric(mean))
```

## Plot 95% confidence intervals by Range for each genetic metric
```{r}
#Allelic Richness Plot
AR_plot <- ggplot(AR_data, aes(x = range, y = mean)) + 
  geom_point(size = 3, show.legend = F) + 
  geom_errorbar(aes(ymin = mean - error, ymax = mean + error, width = 0.3), show.legend = F) + 
  theme_set(theme_cowplot(12)) +
  labs(x = "Range", y = "Mean AR") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20)) + 
  theme(axis.title.x = element_blank())

#Private Allele Frequency Plot
AP_plot <- ggplot(AP_data, aes(x = range, y = mean)) + 
  geom_point(size = 3, show.legend = F) + 
  geom_errorbar(aes(ymin = mean - error, ymax = mean + error, width = 0.3), show.legend = F) +
  theme_set(theme_cowplot(12)) +
  labs(x = "Range", y = "Mean AP") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20)) + 
  theme(axis.title.x = element_blank())

#Common Allele Frequency Plot
A95_plot <- ggplot(A95_data, aes(x = range, y = mean)) + 
  geom_point(size = 3, show.legend = F) + 
  geom_errorbar(aes(ymin = mean - error, ymax = mean + error, width = 0.3), show.legend = F) + 
  theme_set(theme_cowplot(12)) +
  labs(x = "Range", y = "Mean A95") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20)) + 
  theme(axis.title.x = element_blank())

#Observed Heterozygosity Plot
HO_plot <- ggplot(HO_data, aes(x = range, y = mean)) + 
  geom_point(size = 3, show.legend = F) + 
  geom_errorbar(aes(ymin = mean - error, ymax = mean + error, width = 0.3), show.legend = F) + 
  theme_set(theme_cowplot(12)) +
  labs(x = "Range", y = "Mean HO") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20)) + 
  theme(axis.title.x = element_blank())

#Expected Heterozygosity Plot
HE_plot <- ggplot(HE_data, aes(x = range, y = mean)) + 
  geom_point(size = 3, show.legend = F) + 
  geom_errorbar(aes(ymin = mean - error, ymax = mean + error, width = 0.3), show.legend = F) + 
  theme_set(theme_cowplot(12)) +
  labs(x = "Range", y = "Mean HE") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20)) + 
  theme(axis.title.x = element_blank())

#Combine all 95% confidence interval plots 
pdf("/Users/joegunn/Desktop/Grad_School_Stuff/Research/Projects/Smallmouth_Bass_Microsatellites/visualization/genetic_diversity_figures/95_confidence_intervals.pdf", width=12, height=8) 

plot_grid(HO_plot, HE_plot, AR_plot, A95_plot, AP_plot, nrow = 2, ncol = 3, labels = c("A", "B", "C", "D", "E"), label_size = 25)

dev.off()
```
